---
title: "ggplot and maybe Xpose"
author: "Irene Vuu"
date: "7/10/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(xpose4)
```

## getting started

- install.packages("ggplot2")
- install.packages("gridExtra")
- library(ggplot2)
- library(gridExtra)
- library(dplyr)

## ggplot language

- your data are mapped to various aesthesics in a plot such as color, size, or position
- geometric objects are what we see in the plot: the points/lines, scales, bars
- think of plots as:
data + stat transformation + geometric objects + scaling + labels

```{r Theoph, echo = FALSE, warning=FALSE}
df <- data.frame(Theoph)
wt <- df %>%
  group_by(df$Subject) %>%
  summarise(WT=mean(Wt))
  
a <- ggplot(wt, aes(WT)) + geom_histogram()
b <- ggplot(Theoph, aes(Time, conc, color=Subject)) + geom_line()
c <- ggplot(Theoph, aes(Dose, Wt)) + geom_point(aes(color=Subject))
```

## histogram example
```{r a, echo=FALSE, warning=FALSE}
a
```

## spaghetti example
```{r b, echo=FALSE}
b
```

## dotplot example
```{r c, echo=FALSE}
c
```

## things needed for a plot

1. a default dataset and set of mappings from variables to aesthetics,
2. one or more layers, with each layer having one geometric object, one statistical transformation,
3. one position adjustment, and optionally, one dataset and set of aesthetic mappings,
4. one scale for each aesthetic mapping used,
5. a coordinate system,
6. the facet specification

## lets start with the data component
```{r theo}
names(Theoph)
```
- df <- data.frame(Theoph)
- ggplot(df,aes(Time,conc))
```{r theoph df}
ggplot(df,aes(Time,conc))
```

## adding on a geometric object
- ggplot(df,aes(Time,conc)) + geom_point()
```{r theoph point}
ggplot(df,aes(Time,conc)) + geom_point()
```

## aesthetics with geom_point
- alpha(transparency), shape, color, size
- ggplot(df,aes(Time,conc)) + geom_point(alpha=0.2,size=5, color="blue")
```{r theoph fancy point}
ggplot(df,aes(Time,conc)) + geom_point(alpha=0.2, size=5, color="blue")
```

## or you could map to a variable
- ggplot(df,aes(Time,conc)) + geom_point(alpha=0.5, size=1, aes(color=Subject))
- ggplot(df,aes(Time,conc)) + geom_point(alpha=0.7, size=3, aes(color=Wt))
```{r theoph fancy point2}
d <- ggplot(df,aes(Time,conc)) + geom_point(alpha=0.5, size=1, aes(color=Subject))
e <- ggplot(df,aes(Time,conc)) + geom_point(alpha=0.7, size=3, aes(color=Wt))
grid.arrange(d,e,ncol=2)
```

## setting vs mapping
- setting it is what we did first: geom_point(color="blue") 
- vs geom_point(aes(color=Subject))
- aes maps something to a variable in your dataset

## can map in ggplot object too
ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line()
```{r in object}
ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line()
```

## grouping
The "group" argument allows you to map a group of factors
ggplot(df,aes(Time,conc,group=Subject)) + geom_point(aes(size=Wt)) + geom_line(aes(color=Subject))
```{r in object2}
f <- ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line()
g <- ggplot(df,aes(Time,conc,group=Subject)) + geom_point(aes(size=Wt),alpha=0.3) + geom_line(aes(color=Subject))
grid.arrange(f,g,ncol=2)
```

## other good-to-knows
- geom_bar: stat, linetype, fill (bar inside), color (bar outline), alpha, position = "fill/dodge"
- geom_line: stat, linetype, color, alpha
- geom_polygon: adds polygon, good for mapping
- geom_text: adds labels to specific points
- geom_smooth: adds a smoothed conditional means. default is loess if <1000 observations, gam otherwise
- geom_tile: draws rectangles, good for heat maps

## logarithmic scaling
ggplot(df,aes(Time,conc,group=Subject)) + geom_point() + geom_line() + scale_y_log10()
```{r log, warning=FALSE}
ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line() + scale_y_log10()
```

## faceting
adding facet_grid(.~Subject) if you want it separated by columns
adding facet_grid(Subject~.) if you want it separated by rows
```{r facet}
h <- ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line() + facet_grid(.~Subject)
i <- ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line() + facet_grid(Subject~.)
grid.arrange(h,i,ncol=2)
```

## facet wrap
adding facet_wrap(~Subject,ncol/nrow=x) likely better for larger n's
```{r facet2}
j <- ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line() + facet_wrap(~Subject,ncol=3)
k <- ggplot(df,aes(Time,conc,color=Subject)) + geom_point() + geom_line() + facet_wrap(~Subject,nrow=3)
grid.arrange(j,k,ncol=2)
```

## adding a second facet layer
```{r facet layer, echo=TRUE}
df2 <- df %>%
  mutate(Category = if_else(Wt <= 65,
                            "Less than or equal to 65 kg",
                            "Greater than 65 kg"))
ggplot(df2,aes(Time,conc))+geom_line()+facet_grid(Category~Subject)
```

## geom_smooth()
by default will give you a non-parametric, Loess curve including standard error
```{r smooth}
l <- ggplot(df,aes(Time,conc))+geom_point()+geom_smooth(se=TRUE,method="loess")
m <-ggplot(df,aes(Time,conc))+geom_point()+geom_line(aes(group=Subject),alpha=0.3)+geom_smooth(se=TRUE,method="loess")
grid.arrange(l,m,nrow=2)
```

## other geom_smooth options
Use span to control the "wiggliness" of the default loess smoother. The span is the fraction of points used to fit each local regression: small numbers make a wigglier curve, larger numbers make a smoother curve.
```{r smooth2, warning=FALSE}
n <- ggplot(df,aes(Time,conc))+geom_point()+geom_smooth(span=5, method="loess", se=FALSE)
o <- ggplot(df,aes(Time,conc))+geom_point()+geom_smooth(span=0.3, method="loess", se=FALSE)
p <- ggplot(df,aes(Time,conc))+geom_point()+geom_smooth(span=0.15, method="loess", se=FALSE)
grid.arrange(n,o,p, nrow=3)
```

## other geom_smooth options, part 2
you can also use method="lm" for a linear regression (option to use a formula) OR employ a binomial function by using method="glm",method.args=list(family="binomial")
```{r smooth3, warning=FALSE}
df3 <- df %>%
  mutate(Heavy = if_else(Wt <= 65,0,1))
q <- ggplot(df3,aes(Time, conc))+geom_point()+geom_smooth(method="lm", se=FALSE, aes(group=Heavy, color=as.factor(Heavy)))
r <- ggplot(df3,aes(Dose,Heavy))+geom_point()+geom_smooth(method = "glm", method.args = list(family = "binomial"), se=FALSE)
grid.arrange(q,r, nrow=2)
```

## trendplot with standard deviations
using geom_ribbon() and geom_errorbar() to give you standard deviations
```{r trend}
avg <- df3 %>%
  dplyr::group_by(Time) %>%
  dplyr::summarise(AvgConc = mean(conc),
                   SD=sd(conc))

s <- ggplot(filter(avg, !is.na(SD)),aes(Time,AvgConc))+
  geom_point()+
  geom_line()+
  geom_ribbon(aes(ymin=AvgConc-SD,ymax=AvgConc+SD),alpha=0.2)

t <- ggplot(filter(avg, !is.na(SD)),aes(Time,AvgConc))+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=AvgConc-SD,ymax=AvgConc+SD),
                linetype=2,
                size=0.5,
                width=0.2)
grid.arrange(s,t,nrow=2)
```

## multiple trendlines
scale_y_continuous(name= "Primary Y", sec.axis = sec_axis(trans=~. XXX, name="Secondary Y"))
```{r multiple, echo=TRUE, warning=FALSE}
ggplot(df3, aes(Time,conc))+
  geom_smooth(aes(colour="Central"),method="loess",se=FALSE)+
  geom_smooth(aes(Time,y=(conc/10),colour="Gut"),method="loess",se=FALSE)+
  scale_y_continuous(name="Central Concentration",sec.axis = sec_axis(trans = ~.*1, name = "Gut Concentration"))
```

## formatting
```{r cars, echo=TRUE}
cars <- data.frame(mtcars)
ggplot(cars, aes(mpg, wt))+
  geom_point()+
  labs(title="Weight vs MPG",
       subtitle="Correlation plot",
       x="Miles per gallon",
       y="Weight")
```

## more formatting tricks
```{r more cars}
k <- ggplot(cars, aes(mpg, wt, color=as.factor(cyl)))+
  geom_point()+
  labs(title="Weight vs MPG",
       subtitle="Correlation plot",
       x="Miles per gallon",
       y="Weight")

l <- ggplot(cars, aes(mpg, wt, color=as.factor(cyl)))+
  geom_point()+
  labs(title="Weight vs MPG",
       subtitle="Correlation plot",
       x="Miles per gallon",
       y="Weight")+
  scale_colour_manual(values=c("red", "blue", "black"),
                      name="Cylinders",
                      breaks=c("4", "6", "8"),
                      labels=c("4 Cyl", "6 Cyl", "8 Cyl"))

m <- ggplot(cars, aes(mpg, wt, shape=as.factor(cyl)))+
  geom_point()+
  labs(title="Weight vs MPG",
       subtitle="Correlation plot",
       x="Miles per gallon",
       y="Weight")+
  scale_shape_manual(values=c(2, 4, 6),
                      name="Cylinders",
                      breaks=c("4", "6", "8"),
                      labels=c("4 Cyl", "6 Cyl", "8 Cyl"))

n <- ggplot(cars, aes(mpg, wt, shape=as.factor(cyl)))+
  geom_point()+
  labs(title="Weight vs MPG"~(Sigma*mu),
       subtitle="Correlation plot",
       x="Miles per gallon"~(Sigma),
       y="Weight"~(mu))+
  scale_shape_manual(values=c(2,4,6),
                     name="Cylinders"~(delta),
                     breaks=c("4", "6", "8"),
                     labels=c("4 Cyl","6 Cyl","8 Cyl"))
grid.arrange(k,l,m,n,ncol=2)
```

## formatting your text
using theme(), you can format your text in your entire figure 
```{r theme}
ggplot(cars, aes(mpg, wt, color=as.factor(cyl)))+
  geom_point()+
  facet_grid(.~as.factor(cyl))+
  labs(title="Weight vs MPG",
       x="Miles per gallon",
       y="Weight")+
  scale_colour_manual(values=c("red", "blue", "black"),
                      name="Cylinders",
                      breaks=c("4", "6", "8"),
                      labels=c("4 Cyl", "6 Cyl", "8 Cyl")) +
  theme(axis.title.x=element_text(size=15,face="bold"),
        axis.title.y=element_text(size=15,face="bold"),
        axis.text = element_text(size=10,angle = 90),
        axis.ticks = element_line(size = 1),
        axis.ticks.length = unit(0.25, "cm"),
        strip.text.x = element_text(size=20,face="italic"),
        legend.position = c(0.95,0.7),
        legend.title=element_text(size=10),
        legend.text=element_text(size=8),
        panel.spacing = unit(1, "lines"))
```

## special themes
```{r themes}
o <- ggplot(cars, aes(mpg, wt, color=as.factor(cyl)))+
  geom_point()+
  theme_bw()
p <- ggplot(cars, aes(mpg, wt, color=as.factor(cyl)))+
  geom_point()+
  theme_classic()
q <- ggplot(cars, aes(mpg, wt, color=as.factor(cyl)))+
  geom_point()+
  theme_grey()
r <- ggplot(cars, aes(mpg, wt, color=as.factor(cyl)))+
  geom_point()+
  theme_light()
zz <- grid.arrange(o,p,q,r,ncol=2)
```

## exporting your plots
use tiff(), bmp(), jpeg(), and png()
```{r export, echo=TRUE, eval=FALSE}
jpeg("test.jpg", width = 9, height = 8, units = 'in', res = 300)
zz
dev.off()
```

## Xpose4
```{r install, echo=TRUE, warning=FALSE}
#install.packages("xpose4")
library(xpose4)
#library(gridExtra)
```

##building a database
- If your run number is 001, all your NONMEM output files will end in 001.
- xpose.data(runno='001')
- we're going to use the built-in database "simpraz.xpdb"
```{r database, echo=TRUE}
xpdb <- simpraz.xpdb
print(xpdb)
```

## structure
str()
```{r struct, warning=FALSE}
str(xpdb)
```

## names
db.names() tells you a summary of the specified Xpose object to the R console.
```{r dbnames}
db.names(xpdb)
```

## individual plots
ind.plots(). there will be multiple pages, you can specify which page to show with page=...)
```{r indplots, warning=FALSE}
ind.plots(xpdb)
```

## goodness of fit plots
basic.gof(xpdb)
```{r GOF}
basic.gof(xpdb) 
```

## more goodness of fit plots
addit.gof(xpdb)
```{r GOF2}
addit.gof(xpdb) 
```

## DV vs plots
- dv.vs.idv: dv vs independent var
- dv.vs.ipred
- dv.vs.ipred.by.cov: dv vs ipred conditioned on cov's
- dv.vs.ipred.by.idv
- dv.vs.pred
- dv.vs.pred.by.cov
- dv.vs.pred.by.idv
- dv.vs.pred.ipred

## DV vs IPRED conditioned on time
```{r test}
dv.vs.ipred.by.idv(xpdb)
```

## DV vs IPRED and PRED
dv.pres.vs.idv()
```{r dvipredpred}
dv.preds.vs.idv(xpdb)
```

## CWRES plots
- cwres.dist.hist
- cwres.dist.qq
- cwres.vs.cov
- cwres.vs.idv
- cwres.vs.idv.bw
- cwres.vs.pred
- cwres.vs.pred.bw

## CWRES vs pred
dv.pres.vs.idv()
```{r cwres}
cwres.vs.pred(xpdb)
```

## covariate distributions
cov.hist()
```{r histos1}
cov.hist(xpdb)
```

## parameter distributions
parm.hist()
```{r histos2}
parm.hist(xpdb)
```

## eta distributions
ranpar.hist()
```{r histos3}
ranpar.hist(xpdb)
```

## covariates on etas
ranpar.vs.cov()
```{r coveta}
ranpar.vs.cov(xpdb)
```

## VPCs
- you can only do them if you're:
1. Using an xpdb containing a simulation and an estimation problem
2. Using a PsN generated VPC folder
- refer to xpose4 manual

## comparing models
- if you have two (nested?) models to compare, basic.model.comp()
- This creates a stack of four plots, comparing PRED, IPRED, WRES (or CWRES), and IWRES estimates for the two specified model fits.
- dOFV.vs.id(): most/least influential persons b/w 2 models
- dOFV.vs.cov(): dOFV change from cov's b/w 2 models
- dOFV1.vs.dOFV2(): dOFV change b/w 2 models