---
title: "Data Structure, Data Manipulation With Base R Function and Dplyr"
author: "Shen Cheng & Ya-Feng Wen"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```
## Acknowledgement
The contents and strcuture of this workshop refer a lot to R4PMx Workshop in summer 2018 presented by Samuel P Callisto, Ashwin Karanam and David Margraf. 

## Data types in R  
There are multiple ways to represent data in R  

- Numbers  
- Text

## Data types in R

|Data Types|Types of Data Contain|Examples|
|:-:|:-:|:-:|
|character |letters and other text characters           |names, addresses          |
|numeric   |numbers (with decimal points)               |weight, height            |
|integer   |whole numbers                               |age, SSN                  |
|logical   |information on true/false questions         |vegetarian or not         |
|factor    |unordered or ordered categorical information|t-shirt size (S, M, L, XL)|

## Data Types in R 
- Most of the time we want to work with text as a character because they are easier to manipulate.  

- Numbers can also be treated as a character, but this is typically undesirable because we cannot perform any arithmetic manipulations on numbers when they are in this format.

## Converting between common data types in R
- as.logical  
- as.numeric: a <- as.numeric(a)  
- as.character  
- as.factor 

## Data Types in R
```{r}
var <- as.integer(3)
str(var)
```

```{r}
var2 <- var*2
str(var2)
```

```{r}
var.c <- as.character(var)
str(var.c)
```

## Vector
- Vectors contain the same data type.  
- Different with list
```{r}
vec.1 <- 1:5
str(vec.1)
```

```{r}
vec.2 <- c("a", "b", "c", "d", "e")
str(vec.2)
```

```{r}
vec.3 <- c(TRUE,FALSE,F,F,T)
str(vec.3)
```

## data.frame
- A vector of vectors create a matrix or a data.frame.  
- Most of the time you will use data.frames to store your data since each vector can contain a different data type.  
- Matrix: Similar to a data frame, but rows and columns must be the same data type. 

```{r}
df <- data.frame(vec.1, vec.2, vec.3)
names(df) <- c("numbers", "letters", "vowels")
str(df)
```

## data.frame
```{r}
## convert from Factor to character data type
df$letters <- as.character(df$letters)
str(df)
```

## Data Manipulation
- Various formats of original datasets.   
- NONMEM or other pharmacometric softwares have specific requirements for data formats. 

## Set Working Directory  
- Session $\rightarrow$ Set Working Directory $\rightarrow$ Choose Directory   
- base R function: setwd(), getwd()

## Reading Dataset  
- read.csv, readxl, read_sas....  
- read.csv(..., stringAsFactor=TRUE)

## Create New Variables
```{r, eval=FALSE}
midwest$ACT_PCARE_HRS <- midwest$ACT_PCARE/60
Theoph$Time_MIN <- Theoph$Time*60
```

## Make a sub dataset

```{r, Pick rows}
Theoph_ID_1 <- subset(Theoph, Theoph$Subject == 1)
Theoph_ID_2 <- Theoph[Theoph$Subject == 2,]
Theoph_ID_1_3 <- subset(Theoph, Theoph$Subject == 1 | Theoph$Subject == 3)
Theoph_ID_2_4 <- Theoph[Theoph$Subject == 2 | Theoph$Subject == 4,]
```
```{r, Pick columns}
Theoph_Time <- as.data.frame(Theoph$Time)
Theoph_Time <- as.data.frame(Theoph[,4])
Theoph_Time_conc <- as.data.frame(Theoph[,c(4,5)])
```
```{r, Pick rows and columns}
Theoph_ID1_Time_conc <- as.data.frame(Theoph[Theoph$Subject == 1,c(4,5)])
```

## Combine dataset
- cbind  
- rbind  
- merge

## Rename and Reorder Variables 
- colnames(Dataset) <- c()  
- colnames(Dataset)[which(colnames(Dataset) == "a")] <- b  
- Dataset <- Dataset[c(var1, var2, var3,....)]

## Save updated dataset
- write.csv(dataset,'dataset.csv')  
- Save in working directory

## Resources
- Base R Cheat Sheet: https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf  
- The ultimate R Cheat Sheet: https://sites.ualberta.ca/~ahamann/teaching/renr690/R_Cheat_Data.pdf


## Tidyverse

+ A collection of R packages designed for data science <https://www.tidyverse.org/>
+ Includes: readr, tidyr, dplyr, ggplot2, tibble, purr, stringr, forcats, and others
+ Developed by RStudio and it is openly on GitHub
+ Validate R installations are important for commerical use (such as Pharma)
+ Not a replacement for base R
+ Not a finished thing 

## Data Import (readr)

+ Significantly faster
+ More intelligent (dates as dates, and numbers as numbers)
+ Never converts strings to factors
+ Able to import data from SAS, SPSS, and Stata
+ readxl (another package) to import .xls and .xlsx

## Data Format (tidyr)

+ Reshaping and transforming data for further manipulation

## Data Wrangle (dplyr)

+ Pipe (%>%) operator to provide a logical framework for chaining together common data wrangling tasks
+ For subsetting, filtering, summarizing, and generally wrangling data
+ `filter()` select rows
+ `select()` select variables
+ `rename()` rename variables
+ `mutate()` create new variable
+ `arrange()` order rows by variable
+ `group_by()` group rows by variable
+ `summarise()` aggregate the data
+ `case_when` vectorise multiple if_else() statements

## Pipe (%>%)

+ Syntactice sugar: makes code easier to read/write for humans
+ Improving the flow of code
+ Reducing the amount of key strokes needed
+ shortcut: Ctrl + Shift + M (command + shift + M in Mac)

## Examples

+ Example: `mean(diff(data))`

```{r, warning=FALSE, message= FALSE}

data <- c(5,7,11,13)
mean(diff(data))

library(tidyverse) # %>% is not readily available with base R

data %>% 
  diff %>% # calculate the differences between each number
  mean() # calculate the mean
```


## Commonly used base R operators

+ `==` is equal to
+ `>, >=, <, <=` greater than/less than
+ `!=` is not equal to
+ `is.na()` is missing
+ `between()` within range of numbers (inclusive)

## Select variables of interest

+ `starts_with()` select based on prefix
+ `ends_with()` select based on suffix
+ `num_range()` select based on prefix & numeric range
+ `contains()` matches a string within the variable
+ `matches()` more general matching using regular expressions
+ `one_of()` selects columns from a group of names
+ Drop variables: place a minus sign in front of the variable you want to remove `select(df, -Wt)`
+ Rearrange variables: place the `everything()` helper function will fill in the remaining variables you do not mention `select(df, Time, Subject, everything())`
+ Move a variable to the end of the data set: subtract then add it back `select(df, -Wt, weight=Wt)`

## Summarizing data

Any function that returns a single value can be used in summarize().

+ `mean()`
+ `median()`
+ `sd()`
+ `min()`, `max()`
+ `cor()`
+ `n()`
+ `n_distinct()`

## Let's Practice 

+ Use dataset called "PK Covariates" in the datasets folder (download from UMN GitHub UMN_R4PMx_2019)
+ What are the average weight for smoking and non-smoking individuals and how many subjects in each group?
+ How many individuals has genotype 1 in those who are less than 100 kg and equal or greater to 100 kg?
+ What are the Cmax and Tmax for each individual?
+ What are the average Cmax and Tmax based on genotype and race?

## Load the dataset using readr

```{r}
# Load the dataset with readr, change column type to factor and set NA value
covariates <- read_csv("datasets/PK Covariates.csv", 
    col_types = cols(CMT = col_factor(levels = c("1", 
        "2")), FED = col_factor(levels = c("0", 
        "1")), GENO = col_factor(levels = c("0", 
        "1")), MDV = col_factor(levels = c("0", 
        "1")), RACE = col_factor(levels = c("0", 
        "1")), SMOK = col_factor(levels = c("0", 
        "1"))), na = "-99")
```

## Several ways to check missing values

```{r, eval=FALSE}
# searching for every NA and list them, only use for small dataset
is.na(covariates)  
which(is.na(covariates))

# for larger dataset, check if there is any missing value
any(is.na(covariates))  
sum(is.na(covariates))  # how many NAs

complete.cases(covariates) #Find rows with no missing values
#Subset data, keeping only complete cases
df_onlycompleterow <-df[complete.cases(covariates),]  
# Another way to remove rows with NAs
df_onlycompleterow_2 <- na.omit(covariates)  
```


## Average weight (Method 1)
```{r}
# Average weight for smoking and non-smoking individuals
covariates %>% 
  group_by(SMOK) %>% 
  summarise(mean_wt = mean(WT, na.rm = T), n = n_distinct(ID)) # exclude NA in calculation
```

## Average weight (Method 2)
```{r}
## Filter out missing WT record
covariates %>% 
  filter(!is.na(WT)) %>% 
  group_by(SMOK) %>% 
  summarise(mean_wt = mean(WT), n = n_distinct(ID))
```

## Count of individuals has genotype 1 in those who are less than 100 kg and equal or greater to 100 kg
```{r}
covariates %>% 
  mutate(wtLess100 = WT <100) %>% 
  group_by(wtLess100) %>% 
  summarise(genoOneCount = n_distinct(GENO))
```

## Cmax and Tmax for each individual

```{r}
covariates <- covariates %>% 
  group_by(ID) %>% 
  mutate(Cmax = max(CONC),
         temp = case_when(CONC == Cmax ~ TIME),
         Tmax = max(temp, na.rm = T)) %>%
  select(-temp)
head(covariates[,c(2,3,5,13,14)],n=5)
```

## Average Cmax and Tmax based on genotype and race
```{r}
covariates %>% 
  group_by(GENO, RACE) %>% 
  summarise(mean_Cmax = mean(Cmax, na.rm = T), 
            mean_Tmax = mean(Tmax, na.rm = T))
```




















